##Traits associated with the top 20 differentially methylated genes for each group

Author: Samantha Schaffner
Date: November 2, 2020

To supplement gene ontology enrichment and look for top genes of interest to plot, I took the top 20 DM/DHM probes from each comparison. Functions of the genes for these probes were annotated using GeneCards, GWAS Catalog, and EWAS Atlas databases, and this information was placed into supplementary tables.

TUBA8 was one of the top genes with several DM probes in both SNCA-OE and A30P cells after looking at these results, which prompted a figure/pyrosequencing analysis on TUBA8 (see scripts #8 and #8a).

#Libraries
```{r libraries, eval=F}
setwd("~/")
library(dplyr)
```

#Read in data
```{r read in data, eval=F}
DMPs_WT <- read.csv("~/KoborLab/kobor_space/sschaffner/LUHMES_201806/clean_scripts/DMPs_WT_06182020.csv")
DMPs_A30P <- read.csv("~/KoborLab/kobor_space/sschaffner/LUHMES_201806/clean_scripts/DMPs_A30P_06182020.csv")
DHMPs_WT <- read.csv("~/KoborLab/kobor_space/sschaffner/LUHMES_201806/clean_scripts/DHMPs_WT_06182020.csv")
DHMPs_A30P <- read.csv("~/KoborLab/kobor_space/sschaffner/LUHMES_201806/clean_scripts/DHMPs_A30P_06182020.csv")
DMPs_WT_A30P <- read.csv("~/KoborLab/kobor_space/sschaffner/LUHMES_201806/clean_scripts/DMPs_WT_A30P_06182020.csv")
DHMPs_WT_A30P <- read.csv("~/KoborLab/kobor_space/sschaffner/LUHMES_201806/clean_scripts/DHMPs_WT_A30P_06192020.csv")
```

#Subset to unique/overlapping probes: mC
```{r subset probes mC, eval=F}
WT_hits <- DMPs_WT[DMPs_WT$threshold==TRUE,]
A30P_hits <- DMPs_A30P[DMPs_A30P$threshold==TRUE,]
WT_hits_unique <- WT_hits[-which(WT_hits$Name %in% A30P_hits$Name),]
A30P_hits_unique <- A30P_hits[-which(A30P_hits$Name %in% WT_hits$Name),]
shared_mC <- WT_hits[WT_hits$Name %in% A30P_hits$Name,]
shared_A30P <- A30P_hits[A30P_hits$Name %in% WT_hits$Name,]
all.equal(shared_mC$Name, shared_A30P$Name) #false

shared_mC <- shared_mC %>% arrange(Name)
shared_A30P <- shared_A30P %>% arrange(Name)
all.equal(shared_mC$Name, shared_A30P$Name) #TRUE
shared_mC$DB_A30P <- shared_A30P$DB
shared_mC$adjP_A30P <- shared_A30P$adj.P.Val
colnames(shared_mC)[43] <- "DB_WT"

#average the delta beta for shared sites and rank by top 20
shared_mC$avg_DB <- sapply(1:nrow(shared_mC), function(x){
  mean(c(shared_mC$DB_A30P[x], shared_mC$DB_WT[x]))
})
shared_mC <- shared_mC %>% arrange(desc(abs(avg_DB)))
shared_mC_top20 <- shared_mC[1:20,]

#top 20, WT
WT_hits_unique <- WT_hits_unique %>% arrange(desc(abs(DB)))
WT_mC_top20 <- WT_hits_unique[1:20,]

#top 20, A30P
A30P_hits_unique <- A30P_hits_unique %>% arrange(desc(abs(DB)))
A30P_mC_top20 <- A30P_hits_unique[1:20,]
```

#Subset to unique/overlapping probes: hmC
```{r subset probes hmC, eval=F}
WT_hits <- DHMPs_WT[DHMPs_WT$threshold==TRUE,]
A30P_hits <- DHMPs_A30P[DHMPs_A30P$threshold==TRUE,]
WT_hits_unique <- WT_hits[-which(WT_hits$Name %in% A30P_hits$Name),]
A30P_hits_unique <- A30P_hits[-which(A30P_hits$Name %in% WT_hits$Name),]
shared_hmC <- WT_hits[WT_hits$Name %in% A30P_hits$Name,]
shared_A30P <- A30P_hits[A30P_hits$Name %in% WT_hits$Name,]
all.equal(shared_hmC$Name, shared_A30P$Name) #false

shared_hmC <- shared_hmC %>% arrange(Name)
shared_A30P <- shared_A30P %>% arrange(Name)
all.equal(shared_hmC$Name, shared_A30P$Name) #TRUE
shared_hmC$DB_A30P <- shared_A30P$DB
shared_hmC$adjP_A30P <- shared_A30P$adj.P.Val
colnames(shared_hmC)[43] <- "DB_WT"

#average the delta beta for shared sites and rank by top 20
shared_hmC$avg_DB <- sapply(1:nrow(shared_hmC), function(x){
  mean(c(shared_hmC$DB_A30P[x], shared_hmC$DB_WT[x]))
})
shared_hmC <- shared_hmC %>% arrange(desc(abs(avg_DB)))
shared_hmC_top20 <- shared_hmC[1:20,]

#top 20, WT
WT_hits_unique <- WT_hits_unique %>% arrange(desc(abs(DB)))
WT_hmC_top20 <- WT_hits_unique[1:20,]

#top 20, A30P
A30P_hits_unique <- A30P_hits_unique %>% arrange(desc(abs(DB)))
A30P_hmC_top20 <- A30P_hits_unique[1:20,]

#WT/A30P mC annotation in WT mC hits
WT_A30P_mC_sub <- DMPs_WT_A30P[match(WT_mC_top20$Name, DMPs_WT_A30P$Name),]
all.equal(WT_A30P_mC_sub$Name, WT_mC_top20$Name) #TRUE
WT_mC_top20$DB_WT_A30P <- WT_A30P_mC_sub$DB
WT_mC_top20$adjP_WT_A30P <- WT_A30P_mC_sub$adj.P.Val

#WT/A30P mC annotation in A30P mC hits
WT_A30P_mC_sub <- DMPs_WT_A30P[match(A30P_mC_top20$Name, DMPs_WT_A30P$Name),]
all.equal(WT_A30P_mC_sub$Name, A30P_mC_top20$Name) #TRUE
A30P_mC_top20$DB_WT_A30P <- WT_A30P_mC_sub$DB
A30P_mC_top20$adjP_WT_A30P <- WT_A30P_mC_sub$adj.P.Val

#WT/A30P mC annotation in shared mC hits
WT_A30P_mC_sub <- DMPs_WT_A30P[match(shared_mC_top20$Name, DMPs_WT_A30P$Name),]
all.equal(WT_A30P_mC_sub$Name, shared_mC_top20$Name) #TRUE
shared_mC_top20$DB_WT_A30P <- WT_A30P_mC_sub$DB
shared_mC_top20$adjP_WT_A30P <- WT_A30P_mC_sub$adj.P.Val

#WT/A30P hmC annotation in WT hmC hits
WT_A30P_hmC_sub <- DHMPs_WT_A30P[match(WT_hmC_top20$Name, DHMPs_WT_A30P$Name),]
all.equal(WT_A30P_hmC_sub$Name, WT_hmC_top20$Name) #TRUE
WT_hmC_top20$DB_WT_A30P <- WT_A30P_hmC_sub$DB
WT_hmC_top20$adjP_WT_A30P <- WT_A30P_hmC_sub$adj.P.Val

#WT/A30P hmC annotation in A30P hmC hits
WT_A30P_hmC_sub <- DHMPs_WT_A30P[match(A30P_hmC_top20$Name, DHMPs_WT_A30P$Name),]
all.equal(WT_A30P_hmC_sub$Name, A30P_hmC_top20$Name) #TRUE
A30P_hmC_top20$DB_WT_A30P <- WT_A30P_hmC_sub$DB
A30P_hmC_top20$adjP_WT_A30P <- WT_A30P_hmC_sub$adj.P.Val

#WT/A30P hmC annotation in shared hmC hits
WT_A30P_hmC_sub <- DHMPs_WT_A30P[match(shared_hmC_top20$Name, DHMPs_WT_A30P$Name),]
all.equal(WT_A30P_hmC_sub$Name, shared_hmC_top20$Name) #TRUE
shared_hmC_top20$DB_WT_A30P <- WT_A30P_hmC_sub$DB
shared_hmC_top20$adjP_WT_A30P <- WT_A30P_hmC_sub$adj.P.Val
```

#Save
```{r save, eval=F}
write.csv(WT_mC_top20, file="~/KoborLab/kobor_space/sschaffner/LUHMES_201806/clean_scripts/DMPs_WT_top20.csv")
write.csv(A30P_mC_top20, file="~/KoborLab/kobor_space/sschaffner/LUHMES_201806/clean_scripts/DMPs_A30P_top20.csv")
write.csv(shared_mC_top20, file="~/KoborLab/kobor_space/sschaffner/LUHMES_201806/clean_scripts/DMPs_shared_top20.csv")
write.csv(WT_hmC_top20, file="~/KoborLab/kobor_space/sschaffner/LUHMES_201806/clean_scripts/DHMPs_WT_top20.csv")
write.csv(A30P_hmC_top20, file="~/KoborLab/kobor_space/sschaffner/LUHMES_201806/clean_scripts/DHMPs_A30P_top20.csv")
write.csv(shared_hmC_top20, file="~/KoborLab/kobor_space/sschaffner/LUHMES_201806/clean_scripts/DHMPs_shared_top20.csv")
```